
# Assignment 1 Reproducible Research  

Global options
```{r setoptions, echo = TRUE}
library(knitr)
opts_chunk$set(echo = TRUE)
```

##Loading and preprocessing the data  
```{r loader}  
options(scipen = 999) ## prevent exponiential notation in output
data <- read.csv("activity.csv", header = T, sep = ",")  
``` 

##What is the mean total number of steps taken per day?
Lets understand the mean total of the steps taken per day. Ignore missing values.  
```{r FindTotalsperday}  
total.per.day <- tapply(data$steps, data$date, sum, na.rm = TRUE)  
```  

A histgram will help us understand the trend better  
```{r TotalsHistogram}  
hist(total.per.day,     main="Histogram for Steps taken",     xlab="Total steps", border="blue",     col="green" , breaks = 10)  
```  

Finding out the mean and median of the total number of steps taken per day
```{r CalculateStats}
mean.total.per.day <- round(mean(total.per.day))
median.total.per.day <- round(median(total.per.day))
```

The rounded values of mean is `r mean.total.per.day` and the median is `r median.total.per.day`.

##What is the average daily pattern?
Taking a look at the average daily pattern
```{r AvgDailyPattern}
intervals.day <- tapply(data$steps, data$interval, mean, na.rm=T)
plot(intervals.day ~ unique(data$interval), type="l", xlab = "five minute interval", ylab = "mean of steps")
```

Lets see which one is the most active interval of the day
```{r ActiveInterval}
max.pos<- names(intervals.day[which.max(intervals.day)])
```

most active interval of the day is `r max.pos`.

##Imputing missing values
```{r MissingValues}
missing.rows <- nrow(data[!complete.cases(data),])
```

Number of missing rows are `r missing.rows`. 

Replace missing values with mean for intervals
```{r substituteMissing}
data2 <- data  
for (i in 1:nrow(data)){
       if(is.na(data$steps[i])){
              data2$steps[i]<- intervals.day[[as.character(data[i, "interval"])]]
       }
}
```
A new histogram with recalculated values will help us understand the trend better
```{r TotalsHistogram2}
total.per.day2 <- tapply(data2$steps, data2$date, sum, na.rm = TRUE)

hist(total.per.day2,     main="Histogram 2 for Steps taken",     xlab="Total steps", border="blue",     col="green" , breaks = 10)
```

Finding out the mean and median of the total number of steps taken per day with the new data frame without Missing values
```{r CalculateStats2}
mean.total.per.day2 <- round(mean(total.per.day2))
median.total.per.day2 <- round(median(total.per.day2))
```
The new rounded values --> mean is `r mean.total.per.day2` and the median is `r median.total.per.day2`.
The old rounded values were--> mean is `r mean.total.per.day` and the median is `r median.total.per.day`. mean and median have increased. 

##Are there any differences in activity patterns between weekdays and weekends
Create a new column with day type
```{r getDay}
data2$date <- as.Date(data2$date)
#create a vector of weekdays
weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
#convert to factors
data2$wDay <- factor((weekdays(data2$date) %in% weekdays1), 
                     levels=c(FALSE, TRUE), labels=c('weekend', 'weekday'))
```

Plot the weekday Vs weekend plot
```{r PlotLatticePlot}
library(lattice)
plot.data <- aggregate(data2$steps, 
                       by=list(data2$wDay, data2$interval), mean)
names(plot.data) <- c("weekday", "interval", "mean")
xyplot(mean ~ interval |weekday, plot.data, 
       type="l", 
       lwd=1, 
       xlab="Interval", 
       ylab="Number of steps", 
       layout=c(1,2))

```



